Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-02 setup.tex
updating main part of paper

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    \left.\frac{P}{\rho}\right|_{t=0}\right), \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq. \ref{full_energy}) we have
included a thermal relaxation term which restores the disk temperature
$T\propto P/\rho$ to its initial value on a characteristic timescale
$t_c=\beta\Omega_k^{-1}$ where $\beta$ is a dimensionless coefficient
and $\Omega_k=\sqrt{GM_*/r^3}$ is the Keplerian frequency.

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. The equilibrium disk satisfies
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
%where $\Phi_* = -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational
%potential of the central star.
We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H_\mathrm{iso}$ is a characteristic scale-height.  Thus, $\epsilon$
is also the characteristic disk aspect-ratio.

The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $\gamma$ is the constant adiabatic index. We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

The explicit expression for the density field for
$\Gamma\neq1$ is given  by
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right),
\end{align}
Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
$\rho(r,H)=0$. If we parametrize
\begin{align}
  H(r)\equiv h r,
\end{align}
then the imposed sound-speed and disk thickness are related by 
\begin{align}
  h^2(r) = \left[1-\frac{\epsilon^2(r)}{\Gamma-1}\right]^{-2}-1. 
\end{align}
Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define the radial shear rate  
\begin{align}
  S(r,z) \equiv r\frac{\p\Omega}{\p r},  
\end{align}
and the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv 2\Omega(2\Omega + S). 
\end{align}

Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot}, but is also
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr}.\label{vertical_shear}
\end{align}

\subsubsection{Vertically isothermal disks}
For the special case $\Gamma=1$ the disk is isothermal in the
vertical direction. The equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ is then 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks for simplicity.

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure, and $j\equiv r^2\Omega$ is the
specific angular momentum.

For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we consider the second
criterion. For our disk models,
\begin{align}
  &\frac{\p S}{\p r} = c_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = c_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z}. 
\end{align} 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\Omega_k^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side is
$O(\epsilon^2)$. Hence, instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations.

\section{Linear problem}\label{linear}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ and $\nu$ being
the real frequency and growth rate, respectively.

@@

In order to emphasize the radially-local approximation, we
introduce the local Cartesian co-ordinate $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$ using their expressions in the global disk. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.

% We are interested in the axisymmetric stability of the above system,
% on radial length-scales much smaller than the local radius $r$, but
% may be global in the vertical direction. As such, we work within the
% framework of the \emph{vertically global shearing 
%   box} (VGSB) developed by \cite{mcnally14}. In this approximation
% one sets up a local Cartesian co-ordinate  system $(x,y,z)$ centered
% on a fiducial point $(r,\phi,z)=(r_0,\phi_0,0)$ in the global disk, with $\phi_0(t) =
% \Omega(r_0,0)t$. The local $x,y,z$ axis correspond to the
% global radial, azimuthal and vertical directions.

% In this shearing box the equilibrium density, pressure and
% sound-speed are assumed to depend only on $z$, by evaluating these
% quantities at $r=r_0$ in the global disk. The background rotation
% $\Omega$ is approximated by a linear shear in $x$, but left unaltered
% in $z$. However, the vertical shear $\p_z\Omega$ is also assumed to
% only depend on $z$.

% The axisymmetric VGSB equations read 
% \begin{align}
%   & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
%   \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
%   & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
%   v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
%   2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
%   &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
%     v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
%   & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
%   \nabla\cdot\bm{v} -
%   \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), \label{energy_eq}
% \end{align}
% where $\bm{v}$ is the velocity field relative to the bulk flow
% (i.e. $\bm{v}=\bm{0}$ in equilibrium),  
% subscript $0$ indicates evaluation at $r=r_0$. In the energy equation 
% we have added a thermal relaxation term with timescale
% $t_c=\beta\Omega_k^{-1}$ with $\beta$  being a dimensionless input
% parameter.

% \subsection{Linearization}
% We perturb the system such that
%  For clarity we drop the subscript $0$ on the
% frequencies and shear rates below.

% We can eliminate the perturbed velocities between the linearized
% equations to obtain a pair or ordinary differential equations
% for $W,Q$ as
% \begin{align}
%   &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
%   &=\left(\frac{\ii
%     k_x r}{D}\frac{d\Omega^2}{dz} -
%   \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
%   \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
% &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
%   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
% &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
% \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
% &=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
% \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
% \end{align}
% where
% \begin{align}
%   D \equiv \kappa^2 - \sigma^2.
% \end{align}

>>>>

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq. \ref{full_energy}) the source term
$\Lambda$ represent non-adiabatic effects which we
model below. The gas temperature $T$ is defined through the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

\subsection{Thermal relaxation}
We model non-adiabatic effects by defining $\Lambda$ as a thermal
relaxation term,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu t_c}\left(T - T_0\right),
\end{align}
where $T_0=T_0(r,z)$ is the initial temperature, and $t_c$ is a
thermal relaxation timescale parametrized via $\beta$ such that 
\begin{align}\label{beta_def}
  t_c = \beta \Omega_k^{-1},
\end{align}
where $\Omega_k=\sqrt{GM_*/r^3}$ is the Keplerian frequency.

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, for application to protoplanetary disks in 
\S\ref{application} we will model $\beta$ based on physical thermal
timescales, in which case $\beta$ is determined by the equilibrium
disk properties and the perturbation lengthscales of interest.

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
These equations allows us to choose any mid-plane density 
distribution $\rho_0(r)\equiv \rho(r,0)$. In this work we assume a
power-law profile, 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index.

We also consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. %For $\Gamma=\gamma$ the disk is adiabatically stratified
% in the vertical 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio.  In practice, we consider
thin disks with $\epsilon \ll 1$.

For discussion purposes it is useful to define the vertical buoyancy
frequency $N_z$ 
\begin{align}
  N_z^2 \equiv - c_p^{-1}g_z \frac{\p S}{\p z} = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $c_p$ is the heat capacity at constant pressure, $g_z\equiv
\rho^{-1}\p_zP $ and $S\equiv c_p^{-1}\ln{P^{1/\gamma}/\rho}$ is the
specific entropy. The second equality applies to the disk equilibria
described above. 
We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

%\subsection{Vertical structure}

Solving the equilibrium disk equations, we find the explicit
expression for the density field for 
$\Gamma\neq1$ is given  by
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 + \epsilon^2$ there is a disk surface $H_s$
such that $\rho(r,H_s)=0$. 
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{\epsilon^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define % the radial shear rate  
% \begin{align}
%   S(r,z) \equiv r\frac{\p\Omega}{\p r},  
% \end{align}
% and
the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv \frac{1}{r^3}\frac{\p }{\p
    r}\left(r^4\Omega^2\right). %2\Omega(2\Omega + S). 
\end{align}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot}, but is also
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}

\subsubsection{Vertically isothermal disks}
For $\Gamma=1$ the equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ are given by 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks which is the relevant case for irradiated
disks \citep{chiang97}.

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum. For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we consider the second
criterion. For our disk models,
\begin{align}
  &\frac{\p S}{\p r} = c_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = c_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z}. 
\end{align} 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\Omega_k^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side (RHS) is
$O(\epsilon^2)$. Hence, instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
thermal relaxation, as shown below.

\section{Linear problem}\label{linear}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(r, z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ and $\nu$ being
the real frequency and growth rate, respectively.

@@

In order to emphasize the radially-local approximation, we
introduce the local Cartesian co-ordinate $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk in \S\ref{setup}. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.

% The radially-local approximation above is similar in spirit to the
% shearing box framework \citep{goldreich65} or the generalization
% thereof \citep{mcnally14}. The main difference being the terms
% proportional to $\hat{g}_c$ are neglected in the shearing box.
